\huge UNet-HCRF Integration for EEG Pattern Recognition

### 2024-12-11 manuscript/main.tex
edit methods

%% notations; only define those you use frequently
\newcommand{\EE}{{\mathbb{E}}}
\newcommand{\PP}{{\mathbb{P}}}
\newcommand{\QQ}{{\mathbb{Q}}}
\newcommand{\Fb}{\mathbf{F}}
\newcommand{\Ib}{\mathbf{I}}
\newcommand{\hb}{\mathbf{h}}
\newcommand{\xb}{\mathbf{x}}
\newcommand{\one}{\mathbbm{1}}
\newcommand{\Ocal}{\mathcal{O}}

@@

\title{\huge UNet-CRF Integration for Sequence Labeling on EEG Data}

@@

probabilistic graphical model, UNet, fully connected CRF,

@@

@@

\section{Methods}
%Methods: overview of your proposed method, Intuition on why should it be
%better than the state of the art, and details of models and algorithms that
%you developed.

@@

The classical hidden conditional random field (HCRF) model can be formulated as
\begin{equation}
\begin{split}
\PP(y \vert \xb; \theta) &= \sum_{\hb \in H^N}
\mathbb{P}(y, \hb \vert \xb; \theta) \\
&= \frac{\sum_{\hb \in H^N} \exp\left(\Phi(y, \hb, \xb; \theta)\right)}
{\sum_{y^\prime} \sum_{\hb \in H^N}
\exp\left(\Phi(y^\prime, \hb, \xb; \theta)\right)}.
\end{split}
\end{equation}

Different to the design of most existing HCRF models, which they assign
potentials based on the features predefined from the input data $\Ib$.
%$\PP(y, \hb \vert \xb; \theta)$.
%Here, we consider a generative model that constructing
%$\PP(\hb \vert \xb; \theta)$ through a CRF framework then using an 
%unary potential for , 
Our HCRF replace $I$ with features learned by CNN, i.e. $\Fb(\Ib)$. 
Thus the new potentials become:
\begin{equation*}
\Phi(y, \hb, \xb; \theta) = \Phi(y, \hb, \Fb(\Ib); \theta).
%\PP(y, \hb \vert \xb; \theta) = \PP(y \vert \hb, \xb; \theta)
%\cdot \PP(\hb \vert \xb; \theta).
\end{equation*}
Thus, we design the potential of our HCRF model as.
\begin{equation}
\begin{split}
\Phi(y, \hb, \xb; \theta) &= \underbrace{
\sum_{j \in \nu} \phi(h_j, x_j; \omega)
+ \sum_{i \neq j} \psi(h_i, h_j, x_i, x_j; \eta)}_{
% \propto \log\PP(\hb \vert \xb; \theta)
\textrm{Measures log-likelihood $\log\PP(\hb \vert \xb; \theta)$}} \\
&+ \underbrace{
\sum_{j \in \nu} \varphi(y, h_j, x_j; \delta) + \vartheta(y, \xb; \varpi)}_{
% \propto \log\PP(y | \hb,  \xb; \theta)
\textrm{Measures log-likelihood $\log\PP(y \vert \hb, \xb; \theta)$}}
\end{split}
\end{equation}

{\textbf{Unary Potential}}
$\phi(h_j, x_j; \omega)$ measures the likelihood of the local feature
$x_j$ is assigned as the hidden attention state $h_j$.
We generate the likelihood through a softmax operation on $\Fb(\xb)$
the feature maps of CNN with $(N, C_{\textrm{out}})$ dimensions,
by which the potential of each pixel $x_j$ assigned as each state of
hidden state set $\{0, 1, \dots, 5\}$ is drawn and represented by
a matrix with $(N, 2)$ dimension.

The parameter $\omega$ is learned by our end-to-end CNN-Unet training 
structure.

{\textbf{Unary Potential}}
$\varphi(y, h_j; \delta)$ measures the compatibility between global
class label $y$ and the hidden local state $h_j$. This potential is
parametrized as:
\begin{equation*}
\varphi(y, h_{j}; \delta) = \sum_{a \in Y} \sum_{b \in H} \delta_{a, b}
\cdot \one(y = a) \cdot \one(h_j = b).
\end{equation*}

{\textbf{Binary Potential}}
$\psi(h_i, h_j, x_i, x_j; \eta)$
balances the hidden label compatibility of the neighboring pixels on the 
feature image $\Fb(\Ib)$.
We define this potential through a common contrast-sensitive two-kernal 
potentials~\citep{krahenbuhl2011efficient, chen2022end}:
\begin{equation}
\begin{split}
& \psi(h_i, h_j, x_i, x_j; \eta) = \mu(h_i, h_j) \Bigg[
\omega_1 \exp \bigg(
-\frac{\left\lvert p_i - p_j \right\rvert^2}{2\eta_\alpha^2} \\
&\quad
- \frac{\left\lvert \Fb_i(\Ib) - \Fb_j(\Ib)\right\rvert^2}{2\eta_\beta^2}
\bigg)
+ \omega_2 \exp \left(
- \frac{\left\lvert p_i - p_j \right\rvert^2}{2 \eta_\gamma^2}
\right)
\Bigg]
\end{split}
\end{equation}

@@

@@

The existence of the hidden variable $\hb$ makes our model more interpretable.
However, since We will train our model through maximal likelihood method,
introducing the hidden variable will make precisely computation the summation
$\sum_{\hb \in H^N} \PP(y, \hb \vert \xb; \theta)$ intractable.
To address this issue, we make our computation more feasible through a
mean-field variational inference framework, which enables our model
efficiently estimating the log-likelihood $\log\PP(y \vert \xb; \theta)$.

Firstly, we introduce the mean-field varepsilon family approximation of
$\log\PP(\hb \vert y, \xb; \theta)$ as:
\begin{equation*}
\QQ(\hb) = \prod_{i=1}^N q_i(h_i).
\end{equation*}
Then the evidence lower bound (ELBO) under the mean-field family is given by
\begin{equation}
\begin{split}
& \textrm{ELBO}(\QQ) = \EE_{\QQ(\hb)} \log\PP(y, \hb \vert \xb; \theta)
- \EE_{\QQ(\hb)} \log q(\hb) \\
&\quad = \sum_{i=1}^N \EE_{q_i(h_i)}
\left[ \phi(h_j, x_j; \omega) + \varphi(y, h_j, x_j; \delta) \right] \\
&\quad + \sum_{i=1}^N \sum_{j=1}^N \sum_{l,l^\prime} \one_{\{i \neq j\}} 
\QQ(h_i = l) \QQ(h_j = l^\prime) \mu(l, l^\prime) \\
&\quad \Bigg[ \omega_1 \exp \left(
- \frac{\left\lvert p_i - p_j \right\rvert^2}{2\eta_\alpha^2}
- \frac{\left\lvert \Fb_i(\Ib) - \Fb_j(\Ib) \right\rvert^2}{2\eta_\beta^2}
\right) \\
&\quad + \omega_2 \exp \left(
- \frac{\left\lvert p_i - p_j \right\rvert^2}{2\eta_\gamma^2}
\right) \Bigg] \\
&\quad - \sum_{i=1}^N \EE_{q_i} \log q_i(h_i) + \vartheta(y, \xb; \varpi).
\end{split}
\end{equation}

@@

By the coordinate ascent variational inference(CAVI) algorithm,
we can obtain that fix $q_j(h_j), \forall j \neq i$, the optimal $q_i(h_i)$
that maximizes $\textrm{ELBO}(q_i(h_i))$ is given by
\begin{equation}
\QQ(h_i) \propto \exp\{\EE_{\QQ(\hb_{-i})}
\log\PP(y, h_i, \hb_{-i} \vert \xb; \theta)\}.
\end{equation}
Thus, we can get that
\begin{equation}
\begin{split}
q_i(l) &= \frac{1}{Z_i} \exp \Bigg\{
\phi(h_j, x_j; \omega) + \varphi(y, h_j, x_j; \delta) \\
&+ \sum_{j \neq i} \sum_{l^\prime} q(l^\prime) \mu(l, l^\prime) 
\Bigg[ \omega_1 \exp \big(
- \frac{\left\lvert p_i - p_j \right\rvert^2}{2\eta_\alpha^2}\\
&- \frac{\left\lvert \Fb_i(\Ib) - \Fb_j(\Ib) \right\rvert^2}{2\eta_\beta^2}
\big)
+ \omega_2 \exp \big(
- \frac{\left\lvert p_i - p_j 
\right\rvert^2}{2\eta_\gamma^2}
\big)
\Bigg]
\Bigg\}
\end{split}
\end{equation}

@@

\begin{itemize}
\item {\textbf{Updating Variational Family}} $\{q_i(h_i)\}$:
The computationally expensive part of the fixed point iteration of $q_i(h_i)$
comes from the message passing, i.e. the convolution of $\QQ(\hb)$ and
Gaussian kernal in $\psi(h_i, h_j, x_i, x_j; \eta)$. We need $\Ocal(N^2)$
runtime for $N$ pixels when evaluating this term precisely.
To make the computation more feasible, a truncated gaussian kernal will be
employed, which is only supported on spaces within certain proportion of
the untruncated standard deviation. Then the approximate message passing is
linear in the number of pixels $N$.
\item {\textbf{Computation of the ELBO}}: The summation over the product of
$q_i(h_i)q_j(h_j)$ and the gaussian kernal in $\psi(h_i, h_j, x_i, x_j; \eta)$
also requires a $\Ocal(N^2)$ complexity in time.
Owes to a similar truncated gaussian kernal approxiamtion, the runtime can
be reduced to $\Ocal(N)$.
\end{itemize}

\section{Results}
%Results: Description of your experiments and results and a list of questions
%your experiments are designed to answer.

\section{Conclusion}
%Conclusion: discussion and future work.

>>>>

%% notations; only define those you use frequently
\newcommand{\EE}{{\mathbb{E}}}
\newcommand{\Fb}{\mathbf{F}}
\newcommand{\hb}{\mathbf{h}}
\newcommand{\Hc}{\mathcal{H}}
\newcommand{\Ib}{\mathbf{I}}
\newcommand{\Lc}{\mathcal{L}}
\newcommand{\PP}{{\mathbb{P}}}
\newcommand{\QQ}{{\mathbb{Q}}}
\newcommand{\xb}{\mathbf{x}}
\newcommand{\Oc}{\mathcal{O}}
\newcommand{\one}{\mathbbm{1}}
\newcommand{\Ocal}{\mathcal{O}}
\newcommand{\RR}{{\mathbb{R}}}

\let\oldsubsubsection\subsubsection
\renewcommand{\subsubsection}[1]{\oldsubsubsection{\textbf{#1}}}

@@

\title{\huge UNet-HCRF Integration for EEG Pattern Recognition}

@@

classification, probabilistic graphical model, spectrogram,
variational inference, wavelet transformation

@@

\sx{EEG tasks}

@@

\section{Methods}

We begin by introducing the necessary notations.
Let $\xb = (x_0, x_1, \dots, x_N)^\top$ represent the feature vector of the
input image, where $x_0$ denotes the global feature capturing overall image
characteristics, and $x_i$ ($1 \leq i \leq N$) corresponds to a local feature
extracted from the $i$-th pixel. Here, $N$ denotes the total number of pixels
in the image. Let $y \in Y$ represent the class label associated with the image,
where $Y$ is a finite set containing $\vert Y \vert$ possible class labels.
For any feature $\xb$, a corresponding vector of unobserved variables, referred
to as hidden part states and denoted by $\hb = (h_1, \dots, h_N)^\top$,
is introduced. Each $h_i \in H$ represents the hidden part state associated
with $x_i$ for $i = 1, \dots, N$. Here, $H$ is a finite set of hidden part
states, and $\vert H \vert$ denotes the cardinality of this set.

@@

Following the theory of random fields as outlined
by~\citet{quattoni2004conditional} and~\citet{wang2006hidden},
given the feature $\xb$, its corresponding hidden part states $\hb$,
and the class label $y$, a hidden conditional random field (HCRF) can be
expressed in exponential form as follows:
\begin{equation}
\label{eq:hcrf}
\PP(y, \hb \mid \xb; \theta)
= \frac{\exp\left(\Phi(y, \hb, \xb; \theta)\right)}
{\sum_{y^\prime \in Y} \sum_{\hb \in H^N}
\exp\left(\Phi(y^\prime, \hb, \xb; \theta)\right)},
\end{equation}
where $\theta$ is the parameter of the model,
$H^N$ denotes the set of all possible hidden part states of $N$ hidden parts,
and $\Phi(y, \hb, \xb; \theta) \in \RR$ refers to a potential function
depending on the feature $\xb$ and parameterized by $\theta$.

From Equation~\ref{eq:hcrf}, the probability of class label $y$ for the given
feature $\xb$ is:
\begin{equation*}
\begin{split}
\PP(y \vert \xb; \theta) &= \sum_{\hb \in H^N} \PP(y, \hb \mid \xb; \theta) \\
&= \frac{\sum_{\hb \in H^N} \exp\left(\Phi(y, \hb, \xb; \theta)\right)}
{\sum_{y^\prime \in Y} \sum_{\hb \in H^N}
\exp\left(\Phi(y^\prime, \hb, \xb; \theta)\right)}.
\end{split}
\end{equation*}

The potential of the HCRF model, $\Phi(y, \hb, \xb; \theta)$, is defined in the
following form:
\begin{align*}
\Phi(y, \hb, \xb; \theta) &= \underbrace{
\sum_{j \in \nu} \phi(h_j, x_j; \omega) +
\sum_{i \neq j} \psi(h_i, h_j, x_i, x_j; \eta)}_{
% \propto \log \PP( \hb \vert \xb; \theta)
\textrm{Measures log-likelihood $\log \PP(\hb \vert \xb; \theta)$}} \\
& + \underbrace{
\sum_{j \in \nu} \varphi(y, h_j, x_j; \delta) + \vartheta(y, x_0; \varpi)}_{
% \propto \log \PP(y | \hb, \xb; \theta)
\textrm{Measures log-likelihood $\log \PP(y \vert \hb, \xb; \theta)$}},
\end{align*}
where the detailed components are described below.

\textbf{Unary potential} $\phi(h_j, x_j; \omega)$ measures the likelihood of
the local feature $x_j$ is assigned as the hidden attention state $h_j$.
The parameter $\omega$ is learned through the end-to-end CNN-UNet training
structure. This likelihood is computed by applying a softmax operation to the
feature maps produced by the CNN.

\textbf{Binary potential} $\psi(h_i, h_j, x_i, x_j; \eta)$ balances the
hidden label compatibility of the neighboring pixels on the feature $\xb$.
The potential is defined through a common contrast-sensitive two-kernal
potentials~\citep{krahenbuhl2011efficient, chen2022end}:
\begin{equation*}
\begin{split}
& \psi(h_i, h_j, x_i, x_j; \eta) = \mu(h_i, h_j) \Bigg[
\omega_1 \exp \bigg(
-\frac{\left\lvert p_i - p_j \right\rvert^2}{2\eta_\alpha^2} \\
&\quad
- \frac{\left\lvert x_i - x_j\right\rvert^2}{2\eta_\beta^2}
\bigg) + \omega_2 \exp \left(
- \frac{\left\lvert p_i - p_j \right\rvert^2}{2 \eta_\gamma^2}
\right)
\Bigg],
\end{split}
\end{equation*}
where $\mu(h_i, h_j)$ is a label compatibility function that describes
the interactive influences between different pairs of classes,
and $p$ refers to the pixel position.
The parameters $\omega_1$ and $\omega_2$ serve as linear combination weights,
controlling the contributions of the first and second kernels, respectively.
And the parameters $\eta_\alpha$, $\eta_\beta$ and $\eta_\gamma$ control the
influence of the corresponding feature spaces.

\textbf{Unary potential} $\varphi(y, h_j; \delta)$ measures the compatibility
between global class label $y$ and the hidden local state $h_j$.
This potential is parametrized as:
\begin{equation*}
\varphi(y, h_{j}; \delta) = \sum_{a \in Y} \sum_{b \in H} \delta_{a, b}
\cdot \one(y = a) \cdot \one(h_j = b).
\end{equation*}
where $\one(\cdot)$ is an indicator function, and $\delta_{a,b}$ denotes the
likelihood of class label $y = a$ containing a joint with hidden state
$h_j = b$, which is learned during training.

\textbf{Global potential} $\vartheta(y, x_0; \varpi)$ measures the likelihood of
the global feature $x_0$ is assigned as the class label $y$, where the parameter
$\varpi$ is learned during training process.

@@

The inclusion of the hidden variable $\hb$ enhances the interpretability of the
model. However, training the model using maximum likelihood estimation
necessitates the calculation of the summation
$\sum_{\hb \in H^N} \PP(y, \hb \mid \xb; \theta)$, and the presence of the
hidden variable renders the exact computation of this summation intractable.
To circumvent this challenge, a mean-field variational inference framework is
employed, enabling efficient estimation of the log-likelihood
$\log \PP(y \mid \xb; \theta)$.

@@

The mean-field varepsilon family approximation of
$\log \PP(\hb \mid y, \xb; \theta)$ is given by:
\begin{equation*}
\QQ(\hb) = \prod_{i=1}^N q_i(h_i),
\end{equation*}
where $q_i(h_i)$ represents the approximate distribution of the individual
hidden variable.

Under the mean-field family, the evidence lower bound (ELBO) is given by
\begin{equation}
\begin{split}
& \Lc(\QQ) = \EE_{\QQ(\hb)} \log\PP(y, \hb \mid \xb; \theta)
- \EE_{\QQ(\hb)} \log q(\hb) \\
&\quad = \sum_{i=1}^N \EE_{q_i(h_i)}
\left[ \phi(h_j, x_j; \omega) + \varphi(y, h_j, x_j; \delta) \right] \\
&\quad + \sum_{i=1}^N \sum_{j=1}^N \sum_{l,l^\prime} \one_{\{i \neq j\}} 
\QQ(h_i = l) \QQ(h_j = l^\prime) \mu(l, l^\prime) \\
&\quad \Bigg[ \omega_1 \exp \left(
- \frac{\left\lvert p_i - p_j \right\rvert^2}{2\eta_\alpha^2}
- \frac{\left\lvert x_i - x_j \right\rvert^2}{2\eta_\beta^2} \right) \\
&\quad + \omega_2 \exp \left(
- \frac{\left\lvert p_i - p_j \right\rvert^2}{2\eta_\gamma^2}
\right) \Bigg] \\
&\quad - \sum_{i=1}^N \EE_{q_i} \log q_i(h_i) + \vartheta(y, x_0; \varpi).
\end{split}
\end{equation}

\begin{figure*}[tbp]
\centering
\includegraphics[width=\textwidth]{workframe}
\caption{\sx{caption}.}
\label{fig:workframe}
\end{figure*}

@@

By employing the coordinate ascent variational inference (CAVI) algorithm,
an efficient method for approximating posterior distributions in probabilistic
models, the optimization process involves iteratively updating each variational
factor while holding others fixed. Specifically, fixing $q_j(h_j)$,
$\forall j \neq i$, the optimal $q_i(h_i)$ that maximizes $\Lc(q_i(h_i))$
is computed as:
\begin{equation*}
\QQ(h_i) \propto \exp\{\EE_{\QQ(\hb_{-i})}
\log\PP(y, h_i, \hb_{-i} \mid \xb; \theta)\}.
\end{equation*}
Thus, we can get that
\begin{equation*}
\begin{split}
q_i(l) &= \frac{1}{Z_i} \exp \Bigg\{
\phi(h_j, x_j; \omega) + \varphi(y, h_j, x_j; \delta) \\
&+ \sum_{j \neq i} \sum_{l^\prime} q(l^\prime) \mu(l, l^\prime) 
\Bigg[ \omega_1 \exp \bigg(
- \frac{\left\lvert p_i - p_j \right\rvert^2}{2\eta_\alpha^2}\\
&- \frac{\left\lvert x_i - x_j \right\rvert^2}{2\eta_\beta^2}
\bigg)
+ \omega_2 \exp \left(
- \frac{\left\lvert p_i - p_j 
\right\rvert^2}{2\eta_\gamma^2}
\right)
\Bigg]
\Bigg\},
\end{split}
\end{equation*}
where $Z_i$ is the normalizing constant.

@@

\paragraph{Updating variational family}

The computationally intensive step in the fixed-point iteration of $q_i(h_i)$
is the message passing process, which involves convolving $\QQ(\hb)$ with a
Gaussian kernel within the binary potential $\psi(h_i, h_j, x_i, x_j; \eta)$.
This operation requires $\Oc(N^2)$ runtime for $N$ pixels when evaluated
exactly. To improve computational efficiency, a truncated Gaussian kernel is
utilized, which is supported only within a specific proportion of the
untruncated standard deviation. This approximation reduces the complexity of 
message passing to linear time, $\Oc(N)$.

\paragraph{Computation of the ELBO}

The summation over the product of $q_i(h_i)q_j(h_j)$ and the Gaussian kernel
in $\psi(h_i, h_j, x_i, x_j; \eta)$ also incurs a time complexity of $\Oc(N^2)$.
By applying a similar truncated Gaussian kernel approximation, the runtime can
be reduced to $\Oc(N)$.

\section{Results}

\section{Conclusion}
